---
title: "HW6_Spotify_Songs"
---

In this assignment, you will be working with a dataset of songs played on Spotify in 2020, and you'll be writing code to explore the range of danceability scores in the data, using standard deviation and histograms. You'll also write up your findings.

### Task 1: Load the tidyverse

```{r}
library(tidyverse)
```

### Task 2: Load the Spotify Songs dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/spotify_2020.csv

```{r}
spotify_songs <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/spotify_2020.csv")
```

### Task 3: Calculate the mean and standard deviation of the danceability scores

Write code to calculate the mean and standard deviation of the danceability scores in the Spotify dataset. Display those.

```{r}
spotify_songs %>% summarize(mean = mean(danceability), sd = sd(danceability))
```

### Task 4: Create a histogram of the danceability scores

Write code to create a histogram of the danceability scores in the Spotify dataset, based on the example we did in class.

```{r}
# I used a bin width of 0.01 because the bin width of 2 in the example we did in class was made the histogram look like a box
spotify_songs %>% 
  ggplot() +
  geom_histogram(aes(x = danceability), binwidth = 0.01)
```

### Task 5: Add a line to the histogram for the mean danceability score

Take the same code and add a vertical line to the histogram for the mean danceability score, based on the example in class.

```{r}
spotify_songs %>% 
  ggplot() +
  geom_histogram(aes(x = danceability), binwidth = 0.01) +
  geom_vline(aes(xintercept = mean(danceability)), color = "red", linetype = "dashed", size = 1)
```


### Task 6: Add two lines to the histogram to show one standard deviation

Take the previous code and add two lines to the histogram for the 1st standard deviation below and above the mean score.

```{r}
spotify_songs %>% 
  ggplot() +
  geom_histogram(aes(x = danceability), binwidth = 0.01) +
  geom_vline(aes(xintercept = mean(danceability)), color = "red", linetype = "dashed", size = 1) +
  geom_vline(aes(xintercept = mean(danceability) - sd(danceability)), color = "blue", linetype = "dashed", size = 1) +
  geom_vline(aes(xintercept = mean(danceability) + sd(danceability)), color = "blue", linetype = "dashed", size = 1)
```

### Task 7: Describe our results and pitch a story

Write a few sentences about the distribution of danceability scores in the Spotify dataset. Is it a normal distribution, or is it skewed in one direction? What does the mean and one standard distribution tell us about the songs in the dataset? Draft a short story pitch from this data.

This distribution of danceability scores is skewed left, so most of the data is on the right side of the histogram. This means that most of the people viewed the songs in the dataset as more or less "danceable." The mean tells us that, on average, people found the songs in the dataset at least 67% danceable. The standard deviation tells us that a good chunk of the songs (I can't use the empirical rule because it only applies to normal distributions, but I'm guessing the principle still applies) are above 0.50 dancibility. That would make me infer that most people would dance to the songs in the data set.

Story pitch:
I would want to do a story asking people how many of the songs on their playlist they wouldn't want to dance to. As this data suggests, people would more often than not dance to songs than not dance to them. So, I would want to see what songs people wouldn't dance to and why people would say rate them as such. 


### Task 8: Save Your Work and Put it on GitHub

Save this notebook, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.
