---
title: "HW7_MandMs"
---

In this assignment, you'll be picking up where you left off with the M&Ms activity. You'll load the data and perform some basic statistical analysis. Follow the steps below to complete the assignment.

### Task 1: Load the tidyverse
```{r}
library(tidyverse)
```


### Task 2: Load the M&Ms dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/mandms.csv
```{r}
MAM <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/mandms.csv")
```


### Task 3: Sample 5 rows of data from the dataset, and then sample 10 rows, saving each to separate dataframes called sample5 and sample10.
```{r}
sample5 <- MAM %>% sample_n(5)
sample10 <- MAM %>% sample_n(10)
```


### Task 4: Calculate the mean and standard deviation of the samples, using one of the colors.
```{r}
# I will be using the color black
sample5 %>% summarize(mean = mean(black), sd = sd(black))
sample10 %>% summarize(mean = mean(black), sd = sd(black))
```


### Task 5: Calculate the mean and standard deviation of the population, using the same color as Task 4
```{r}
MAM %>% summarize(mean = mean(black), sd = sd(black))
```


### Task 6: Compare the sample and population statistics

Compare the three sets of means and standard deviations. Are they similar?

The means and standard deviations are similar. However, the mean and standard deviation of sample5 are farther away from the mean and standard deviation of the population. Conversely, the mean and standard deviation of sample10 are closer to that of the population. 


### Task 7: Describe the results

Write a few sentences about what you found in the data. What does this tell you about the M&Ms dataset and about the act of random sampling?

The standard deviation of the population number of black M&M's being around 15 tells us that there is some variation in the data and it is not very clustered around the mean. Also, the population mean of black M&M's being greater than 50 (the percentage how many black M&M's there are) means that the average sample of black M&M's contained a higher ratio of Black M&M's than the population.

Additionally, the sample's means and standard deviations being similar to the population mean and standard deviation tells us that random sampling, on average, will produce a sample that is fairly similar in attributes to the population. 

When finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
