---
title: "final_project_submission"
output: html_document
---

########## Alex Taylor's Final Project for JOUR405 ##########

This is my final project submission for JOUR405. The file is named "final_project_submission" because I have condensed and re-written some of the code in my original final project file, "final_project," to make it easier to read.



### Set up

Load necessary packages:
```{r}
library("tidyverse")
library("tidycensus")
knitr::opts_chunk$set(echo = TRUE)
options(scipen = 999)
```

Load data:
```{r}
#load unemployment rate data for October of 2024 from the BLS website
#I complied the data from the BLS website into a google sheet, then uploaded it as a CSV to my repository
county_urate <- read_csv("https://raw.githubusercontent.com/awtaylor17/JOUR405_Spring25/refs/heads/main/project/JOUR405_URateDate_Compiled2.csv")

#load election data provided to me by Derek Willis
election_data <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_pres_county.csv")

#load demographic data from the Census Bureau’s 2023 ACS
Sys.getenv("CENSUS_API_KEY")

#retrieving the amount of people who identified as white alone in each county
md_white_2023 <- get_acs(geography = "county",
              variables = c("B02001_002"),
              state = "MD",
              year = 2023)

#retrieving the amount of estimated income in each county
md_income_2023 <- get_acs(geography = "county",
              variables = c("B07411_001"),
              state = "MD",
              year = 2023)

#retrieving the total population (for race) in each county
md_racetotal_2023 <- get_acs(geography = "county",
              variables = c("B02001_001"),
              state = "MD",
              year = 2023)

#retrieving the median age in each county
md_medianage_2023 <- get_acs(geography = "county",
              variables = c("B01002_001"),
              state = "MD",
              year = 2023)

#retrieving the number of males in each county
md_male_2023 <- get_acs(geography = "county",
              variables = c("B01001_002"),
              state = "MD",
              year = 2023)

#retrieving the total population (for gender) in each county
md_totalgender_2023 <- get_acs(geography = "county",
              variables = c("B01001_001"),
              state = "MD",
              year = 2023)

```

While I could probably assume that md_racetotal_2023 and md_totalgender_2023 have the same values, I am using the separate data frames just to be safe. 

Clean data and make it easier to interpret:
```{r}
#create trump_pct
election_data <- election_data %>% 
  mutate(
    trump_pct = trump/total * 100
  )

#change estimate name to income
md_income_2023 <- md_income_2023 %>% 
  rename(
    income = estimate
  )

#create male_pct
md_male_2023 <- full_join(md_male_2023, md_totalgender_2023, by = "GEOID")

md_male_2023 <- md_male_2023 %>%
  mutate(
    male_pct = estimate.x/estimate.y * 100
  )

#change estimate name to median age
md_medianage_2023 <- md_medianage_2023 %>% 
  rename(
    median_age = estimate
  )

#create white_pct
md_white_2023 <- full_join(md_white_2023, md_racetotal_2023, by = "GEOID")

md_white_2023 <- md_white_2023 %>%
  mutate(
    white_pct = estimate.x/estimate.y * 100
  )

```

Merge data into one data frame to make analysis easier:

```{r}

#select certain values to make the merged datafram less busy
md_income_2023 <- md_income_2023 %>% 
  select(
    GEOID, NAME, income
  )

md_male_2023 <- md_male_2023 %>%
  select(
    GEOID, male_pct
  )

md_medianage_2023 <- md_medianage_2023 %>% 
  select(
    GEOID, median_age
  )

md_white_2023 <- md_white_2023 %>% 
  select(
    GEOID, white_pct
  )

#merge dataframes that have GEOID
all_data <- full_join(md_income_2023, md_male_2023, by = "GEOID")
all_data <- full_join(all_data, md_medianage_2023, by = "GEOID")
all_data <- full_join(all_data, md_white_2023, by = "GEOID")

#rename county values to match for merger
all_data <- all_data %>%
  mutate(NAME = str_replace(NAME, " County, Maryland", "")) |> 
  mutate(NAME = str_replace(NAME, " city, Maryland", " City"))

all_data <- all_data %>%
  mutate(NAME = case_when(
    NAME == "Baltimore" ~ "Baltimore County",
    TRUE ~ NAME
  ))

all_data <- all_data %>%
  mutate(NAME = case_when(
    NAME == "St. Mary's" ~ "Saint Mary's",
    TRUE ~ NAME
  ))

#rename column names to match for merger
all_data <- all_data %>% 
  rename(
    County = NAME
  )

election_data <- election_data %>% 
  rename(
    County = jurisdiction
  )

#merge the unemployment rate and election data with all_data
all_data <- full_join(all_data, county_urate, by = "County")
all_data <- full_join(all_data, election_data, by = "County")

```

Note: Based on your suggestion to use mutate() and case_when() to rename values, I Googled the proper format and was shown the structure used above. That is also how I learned about the select() function.


### Analysis

## Descriptive Stats

[ADD INTRO TO DESCRIPTIVE STATS]

```{r}
summary(election_data)
```



[ADD RESULTS OF DESCRIPTIVE STATS]

## Analysis of Variables

# Unemployment rate

[ADD INTRO TO U-RATE]

```{r}
#calculate the correlation
all_data %>% summarize(cor(U_Rate, trump_pct, method = "pearson"))

#visualize the correlation
all_data %>% 
  ggplot() +
  geom_point(aes(x=trump_pct, y=U_Rate)) +
  geom_smooth(aes(x=trump_pct, y=U_Rate), method = "lm") +
  geom_text(aes(x=trump_pct, y=U_Rate, label=County), size = 3)

#calculate linear regression model
summary(lm(`trump_pct` ~ `U_Rate`, data = all_data))

```

[ADD RESULTS OF U-RATE]


# Percent white

[ADD INTRO TO PCT WHITE]

```{r}
#calculate the correlation
all_data %>% summarize(cor(white_pct, trump_pct, method = "pearson"))

#visualize the correlation
all_data %>% 
  ggplot() +
  geom_point(aes(x=trump_pct, y=white_pct)) +
  geom_smooth(aes(x=trump_pct, y=white_pct), method = "lm") +
  geom_text(aes(x=trump_pct, y=white_pct, label=County), size = 3)

#calculate linear regression model
summary(lm(`trump_pct` ~ `white_pct`, data = all_data))

```

[ADD RESULTS OF PCT WHITE]


# Median age

[ADD INTRO TO MEDIAN AGE]

```{r}
#calculate correlation
all_data %>% summarize(cor(median_age, trump_pct, method = "pearson"))

#visualize correlation
all_data %>% 
  ggplot() +
  geom_point(aes(x=trump_pct, y=median_age)) +
  geom_smooth(aes(x=trump_pct, y=median_age), method = "lm") +
  geom_text(aes(x=trump_pct, y=median_age, label=County), size = 3)

#calculate linear regression model
summary(lm(`trump_pct` ~ `median_age`, data = all_data))
```

[ADD RESULTS OF MEDIAN AGE]


# Percent male

[ADD INTRO TO PCT MALE]

```{r}
#calculate correlation
all_data %>% summarize(cor(male_pct, trump_pct, method = "pearson"))

#visualize correlation
all_data %>% 
  ggplot() +
  geom_point(aes(x=trump_pct, y=male_pct)) +
  geom_smooth(aes(x=trump_pct, y=male_pct), method = "lm") +
  geom_text(aes(x=trump_pct, y=male_pct, label=County), size = 3)

#calculate linear regression model
summary(lm(`trump_pct` ~ `male_pct`, data = all_data))

```

[ADD RESULTS OF PCT MALE]


# Income

[ADD INTRO TO INCOME]

```{r}
#calculate correlation
all_data %>% summarize(cor(income, trump_pct, method = "pearson"))

#visualize correlation
all_data %>% 
  ggplot() +
  geom_point(aes(x=trump_pct, y=income)) +
  geom_smooth(aes(x=trump_pct, y=income), method = "lm") +
  geom_text(aes(x=trump_pct, y=income, label=County), size = 3)

#calculate linear regression model
summary(lm(`trump_pct` ~ `income`, data = all_data))

```

[ADD RESULTS OF INCOME]

## Multiple Regression

[ADD INTRO TO MULTIPLE REGRESSION]

```{r}
#create multiple regression models, steadily narrowing down the variables
mr_model_1 <- lm(trump_pct ~ median_age + white_pct + male_pct + income + U_Rate, data = all_data)
summary(mr_model_1)

mr_model_2 <- lm(trump_pct ~ white_pct + male_pct + income + U_Rate, data = all_data)
summary(mr_model_2)

mr_model_3 <- lm(trump_pct ~ white_pct + male_pct + income, data = all_data)
summary(mr_model_3)

mr_model_4 <- lm(trump_pct ~ white_pct + income, data = all_data)
summary(mr_model_4)


#create comparison chart
model_names <- c("Full model", "Model 2", "Model 3", "Model 4")
adj_r_squared <- c(summary(mr_model_1)$adj.r.squared,
                   summary(mr_model_2)$adj.r.squared,
                   summary(mr_model_3)$adj.r.squared,
                   summary(mr_model_4)$adj.r.squared)

model_comparison <- data.frame(Model = model_names, Adjusted_R_squared = adj_r_squared)
ggplot(model_comparison, aes(x = reorder(Model, Adjusted_R_squared), y = Adjusted_R_squared)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Model Comparison by Adjusted R-squared",
       x = "Model")
```


[ADD RESULTS OF MULTIPLE REGRESSION]


### Story Pitch

[ADD STORY PITCH/FINAL STUFF]

